---
title: "4. Using map projections"
author: "Dan Kelley (https://orcid.org/0000-0001-7808-5911)"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    fig_caption: yes
    fig_width: 3.5
    fig_height: 3.5
    dpi: 72
    dev.args: list(pointsize=11)
vignette: >
  %\VignetteIndexEntry{4. Using map projections}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- HOW TO BUILD THE VIGNETTE. -->
<!-- Look in oce.Rmd for instructions. -->

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

**Abstract.** An overview of the handling of map projections in oce is
presented, along with a few examples of common projections.

# Introduction

Map projections provide methods for representing the three-dimensional surface
of the earth as two-dimensional plots.  Although most oceanographers are likely
to be familiar with the basic ideas of map projection, they may find it helpful
to consult the wide literature on this topic, whether to learn about the
details of individual projections and to get advice on the best choice of
projection for a particular task. Snyder (1987), Snyder (1993) and Snyder and
Voxland (1994) are all excellent sources on these topics, and the first and
last of these are available for free online.

Oce handles map projections by calling the `sf_project` function of the
`sf` R package, and so the notation for representing the projection is
borrowed from `sf`.  This system will be familiar to many readers,
because it is used in other R packages, and more widely in the software called
PROJ, which has interfaces in python and other programming languages (PROJ
contributors, 2020). Since oce uses inverse projections in some of its
graphical work, only PROJ projections that have inverses are incorporated into
oce.  Furthermore, some projections are omitted from oce because they have been
witnessed to cause problems on the oce developer's computers, including
infinite loops and core dumps. See the help for the oce function `mapPlot` for
a list of the available projections, and some advice on choosing them.

You must ensure that the `sf` package is installed for the examples provided
in this vignette to work.  If this package is not installed, it may be installed
with
```{r eval=FALSE}
install.packages("sf")
```
(Of course, this step need not be repeated to use `oce`.)


There are far too many projections to illustrate here. See
<https://dankelley.github.io/r/2015/04/03/oce-proj.html> for a blog item that
provides examples of the available projections. Note that some have a problem
with spurious horizontal lines. This can result from coastline segments that
cross the edge of the plotting area, and getting rid of this is tricky enough
to be the heart of the longest-lived bug in the `oce` issue list, i.e.
<https://github.com/dankelley/oce/issues/388>. In some instances the function
`coastlineCut()` can help, but it is provisional and subject to change.


# Avoiding projections for small regions

For map views that span only a few tens or hundreds of kilometers, it may be
sufficient to plot directly, in rectilinear longitude-latitude space, but with
an appropriate aspect ratio so that circular islands will appear circular in
the plot.

To see the need for setting the aspect ratio, consider the following view of
eastern North America.

```{r}
library(oce)
data(coastlineWorld)
lon <- coastlineWorld[["longitude"]]
lat <- coastlineWorld[["latitude"]]
par(mar=c(4, 4, 0.5, 0.5))
plot(lon, lat, type="l",
     xlim=c(-130, -50), ylim=c(40, 50))
```

Readers familiar with this region will notice that the coastline
shapes are distorted.  The solution to this problem is to set
the `asp` argument of `plot()` to a value that will compensate
for meridional convergence at the poles, as below.

```{r}
par(mar=c(4, 4, 0.5, 0.5))
plot(lon, lat, type="l",
     xlim=c(-130, -50), ylim=c(40, 50), asp=1/cos(45*pi/180))
```

Although the above approach is not exactly taxing, the effort of setting the
aspect ratio and setting line-type plots can be spared by using the generic
`plot()` function for `coastline` objects, as follows.

```{r}
plot(coastlineWorld, clongitude=-90, clatitude=45, span=7000)
```

Note that this `plot` is being transferred to the specialized method for a
`coastline` object (defined by the `oce` package) and that this means that
there are some arguments in addition to those of the normal `plot` function.
In this example, the second and third arguments specify the central spot of
interest, and the fourth is a suggested diagonal span, in kilometres.  (The
span is not always obeyed exactly, and the results also depend on aspect ratio
of the plot device.)

The graphs shown above share a common strength: the axes for longitude and
latitude are orthogonal, and the scale along each axis is linear. This makes it
easy for readers to identify the location of features of the diagrams. However,
using linear axes on large-scale views leads to distortions of coastline shapes
and relative feature sizes, and motivates the use of map projections. (Readers
who question the previous sentence are again encouraged to consult some of the
references in the bibliography.)

# Choosing a map projection

## World views

It makes sense to switch to a coarser coastline file for the examples to
follow, because plotting a detailed coastline in a world view leads to a
scribbling effect that obscures the large-scale coastline shapes.

```{r}
data(coastlineWorld)
```

The function used to produce maps with projections is
`mapPlot`, and with default arguments it produces

```{r fig.cap="World coastline with default (Mollweide) projection.", fig.width=5, fig.height=2.7, dpi=72, dev.args=list(pointsize=10)}
par(mar=c(1.5, 1.5, 0.5, 0.5))
mapPlot(coastlineWorld, col="lightgray")
```

Note that the land is coloured in this example, but that sometimes doing that
will yield odd-looking plots, because of a bug in how some projections are
handled in oce.

This default plot uses the Mollweide projection (see any of the Snyder
references in the bibliography), which is set by the default value
`"+proj=moll"` for `mapPlot`.
Another popular world view is the Robinson projection, which has been used by
Rand McNally, the National Geographic Society, and other groups.

**Exercise 1.** Plot the world with a Robinson projection, using gray shading for land.

**Exercise 2.** Plot an image of world topography with the Mollweide projection.

## Polar views

In both these views (and in most world-spanning projections) there is
substantial distortion at high latitudes. The Stereographic projection offers a
solution to this problem, e.g. in the following (which employs a trick on the
latitude limit, specifying an image point on the other side of the planet, with
the pole in between).

```{r fig.cap="Polar view with stereographic projection.", fig.width=5, fig.height=5, dpi=72, dev.args=list(pointsize=10)}
par(mar=c(1.5, 1.5, 0.5, 0.5))
mapPlot(coastlineWorld, longitudelim=c(-130, 50), latitudelim=c(60, 120),
        projection="+proj=stere +lat_0=90", col='gray')
```

Here, the region of interest is defined with `longitudelim` and `latitudelim`,
instead of with `clongitude`, `clatitude` and `span`. Note that the latitude
limit is set to extend past the pole, which is a way of making the plot include
the pole.  See the table in the documentation for `mapPlot` for a listing of
arguments, and for citations to external resources that explain what they mean.

**Exercise 3.** Draw a view of Antarctica and the Southern Ocean.


## Mid-latitude views

The Lambert Conformal view is often used in maps that span wide
longitudinal ranges, e.g. a map of Canada may be produced as follows.

```{r}
par(mar=c(1.5, 1.5, 0.5, 0.5))
mapPlot(coastlineWorld,
    longitudelim=c(-130,-55), latitudelim=c(45, 70),
    projection="+proj=lcc +lat_1=50 +lat_2=65 +lon_0=-100", col='gray')
```

Here, `lat_1` and `lat_2`, the latitudes where the projection cone intersects
the earth, are set to span the populated region of Canada and the northern USA,
and `lon_0`, the meridian that will be vertical on the plot, is chosen to lie
near the middle of the continent.

**Exercise 4.** Plot the eastern North Atlantic using a Mercator projection.

**Exercise 5.** Plot the eastern North Atlantic using an Albers Equal-Area
Conic projection.



# Adding to map plots

Several functions are provided by oce to draw additional features on maps.
These include
`mapText` for adding text,
`mapPoints` for adding points,
`mapLines` for adding lines,
`mapContour` for adding contours, and
`mapImage` for adding images.
Each of these requires a map to have been drawn first with
`mapPlot`. The interfaces ought to be similar
enough to base-R functions that readers can decide how to accomplish
common tasks.

**Exercise 6.** Plot the world with the Goode projection, superimposing sea surface temperature as contours.



# Solutions to exercises

**Exercise 1.** Plot the world with a Robinson projection, using gray shading for land.

```{r fig.cap="World coastline with Robinson projection.", fig.width=5, fig.height=2.7, dpi=72, dev.args=list(pointsize=10)}
par(mar=rep(0.5, 4))
mapPlot(coastlineWorld, col="lightgray", projection="+proj=robin")
```

**Exercise 2.** Plot an image of world topography with the Mollweide projection.

Note that some adjustment of the margins is required to fit the colorbar.

```{r fig.cap="World topography with Mollweide projection.", fig.width=5, fig.height=2.7, dpi=72, dev.args=list(pointsize=10)}
par(mar=c(1.5, 1, 1.5, 1))
data(topoWorld)
topo <- decimate(topoWorld, 2) # coarsen grid: 4X faster plot
lon <- topo[["longitude"]]
lat <- topo[["latitude"]]
z <- topo[["z"]]
cm <- colormap(name="gmt_globe")
drawPalette(colormap=cm)
mapPlot(coastlineWorld, projection="+proj=moll", grid=FALSE, col="lightgray")
mapImage(lon, lat, z, colormap=cm)
```

**Exercise 3.** Draw a view of Antarctica and the Southern Ocean.

```{r}
par(mar=c(1.5, 1.5, 0.5, 0.5))
mapPlot(coastlineWorld, longitudelim=c(-180, 180), latitudelim=c(-130, -50),
       projection="+proj=stere +lat_0=-90", col='gray', grid=15)
```


**Exercise 4.** Plot the eastern North Atlantic using a Mercator projection.

```{r fig.cap="Eastern North Atlantic with Mercator projection.", fig.width=5, fig.height=5, dpi=72, dev.args=list(pointsize=10)}
par(mar=c(1.5, 1.5, 0.5, 0.5))
mapPlot(coastlineWorld, col="lightgray",
        projection="+proj=merc",
        longitudelim=c(-80, -40), latitudelim=c(20, 60))
```

**Exercise 5.** Plot the eastern North Atlantic using an Albers Equal-Area
Conic projection.

In the solution given below, note how the `lon_0` argument is set to be at the
central value of `longitudelim`. This makes the $60^\circ$W line be vertical on
the plot. The settings for `lat_1` and `lat_2` are the locations at which the
cone intersects the earth, so distortion is minimized at those latitudes. Note
how the lines of constant longitude are changed, compared with those in the
Mercator view. It may be informative to compare the relative ratio of the area
of Nova Scotia (at $45^\circ$N) and Ungava Bay (at $60^\circ$N) in this Albers
Equal-Area projection, and the Mercator projection shown in the previous
diagram.



```{r fig.cap="Eastern North Atlantic with Albers equal-area projection.", fig.width=5, fig.height=5, dpi=72, dev.args=list(pointsize=10)}
par(mar=c(1.5, 1.5, 0.5, 0.5))
mapPlot(coastlineWorld, col="lightgray",
        projection="+proj=aea +lat_1=30 +lat_2=50 +lon_0=-60",
        longitudelim=c(-80, -40), latitudelim=c(20, 60))
```

**Exercise 6.** Plot the world with the Goode projection, superimposing sea surface temperature as contours.

Note that the temperature is provided by the `ocedata` package.

```{r fig.cap="SST contours with the Goode projection", fig.width=5, fig.height=2.7, dpi=72, dev.args=list(pointsize=10)}
par(mar=rep(0.5, 4))
mapPlot(coastlineWorld, projection="+proj=goode", col="lightgray")
if (requireNamespace("ocedata", quietly=TRUE)) {
    data(levitus, package="ocedata")
    mapContour(levitus[['longitude']], levitus[['latitude']], levitus[['SST']])
}
```


# References

PROJ contributors. "PROJ Coordinate Transformation Software Library." Open
Source Geospatial Foundation, 2020.
<https://proj.org/>

Snyder, John P. "Map Projections-a Working Manual." Washington: U.S.
Geological survey professional paper 1395, 1987.
<https://pubs.usgs.gov/pp/1395/report.pdf>

Snyder, John Parr. "Flattening the Earth: Two Thousand Years of Map
Projections." Chicago, IL: University of Chicago Press, 1993.
<https://press.uchicago.edu/ucp/books/book/chicago/F/bo3632853.html>

Snyder, John P., and Philip M. Voxland. "An Album of Map Projections."
U. S.  Geological Survey Professional Paper. Washington, DC, USA: U. S.
Department of the Interior: U. S. Geological Survey, 1994.
<https://pubs.er.usgs.gov/publication/pp1453>


