---
title: "Map Projections in Oce"
author: "Dan Kelley (https://orcid.org/0000-0001-7808-5911)"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    fig_caption: yes
    fig_width: 5
    fig_height: 5
    dpi: 72
    dev.args: list(pointsize=11)
vignette: >
  %\VignetteIndexEntry{Map projections in oce}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- HOW TO BUILD THE VIGNETTE. -->
<!-- Look in oce.Rmd for instructions. -->

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

**Abstract.** An overview of the handling of map projections in oce is
presented, along with a few examples of common projections.

# Introduction

Map projections provide methods for representing the three-dimensional surface
of the earth as two-dimensional plots.  Although most oceanographers are likely
to be familiar with the basic ideas of map projection, they may find it helpful
to consult the wide literature on this topic, whether to learn about the
details of individual projections and to get advice on the best choice of
projection for a particular task. Snyder (1987), Snyder (1993) and Snyder and
Voxland (1994) are all excellent sources on these topics, and the first and
last of these are available for free online.

Oce handles map projections by calling the \code{projection} function of the
\code{rgdal} R package, and so the notation for representing the projection is
borrowed from \code{rgdal}.  This system will be familiar to many readers,
because it is used in other R packages, and more widely in the software called
PROJ.4, 2019), which has interfaces in python and other programming languages
(PROJ.4, 2019). Since oce uses inverse projections in some of its graphical
work, only PROJ.4 projections that have inverses are incorporated into oce.
Furthermore, some projections are omitted from oce because they have been
witnessed to cause problems on the oce developer's computers, including
infinite loops and core dumps. See the help for the oce function
[`mapPlot`](../help/mapPlot) for a list of the available projections, and some
advice on choosing them.

# Avoiding projections for limited-area plots

For map views that span only a few tens or hundreds of kilometers, it may be
sufficient to plot directly, in rectilinear longitude-latitude space, but with
an appropriate aspect ratio so that circular islands will appear circular in
the plot.

To see the need for setting the aspect ratio, consider the following view of
eastern North America.
shape.

```{r}
library(oce)
data(coastlineWorldFine, package="ocedata")
lon <- coastlineWorldFine[["longitude"]]
lat <- coastlineWorldFine[["latitude"]]
par(mar=c(4, 4, 0.5, 0.5))
plot(lon, lat, type="l",
     xlim=c(-80, -50), ylim=c(40, 50))
```

Readers familiar with this region will see immediately that the coastline
shapes are distorted.  However, setting `asp` to yield equal scaling of
east-west and north-south *distances* at a central latitude yields the
following.


```{r}
par(mar=c(4, 4, 0.5, 0.5))
plot(lon, lat, type="l",
     xlim=c(-80, -50), ylim=c(40, 50), asp=1/cos(45*pi/180))
```

This scaling yields results that more faithfully represent coastline shapes.
The effort of setting the aspect ratio and of using line-type plots is spared
by using the generic `plot()` function for `coastline` objects, as follows.

```{r}
plot(coastlineWorldFine,
     longitudelim=c(-69, -57), latitudelim=c(43, 47))
```

Irrespective of shape distortion, the graphs shown above share a common
stength: the axes for longitude and latitude are orthogonal, and the scale
along each axis is linear. This makes it easy for readers to identify the
location of features of the diagrams. However, using linear axes on large-scale
views leads to distortions of coastline shapes and relative feature sizes, and
motivates the use of map projections. (Readers who question the previous
sentence are again encouraged to consult some of the references in the
bibliography.)

# Choosing a map projection

## World views

It makes sense to switch to a coarser coastline file for the examples to
follow, because plotting a detailed coastline in a world view leads to a
scribbling effect that obscures the large-scale coastline shapes.

```{r}
data(coastlineWorld)
```

The function used to produce maps with projections is
[`mapPlot`](../help/mapPlot), and with default arguments it produces

```{r fig.height=3}
par(mar=c(1.5, 1.5, 0.5, 0.5))
mapPlot(coastlineWorld)
```

This default plot uses the Mollweide projection (see any of the Snyder
references in the bibliography), which is set by the default value
`"+proj=moll"` for [`mapPlot`](../help/mapPlot). Another popular world view is
the Robinson projection, which has been used by Rand McNally, the National
Geographic Society, and other groups.

**Exercise 1.** Plot the world with a Robinson projection, using gray shading
for land.

## Polar views

In both these views (and in most world-spanning projections) there is
substantial distortion at high latitudes. The Stereographic projection offers a
solution to this problem, e.g. in the following (which employs a trick on the
latitude limit, specifying an image point on the other side of the planet, with
the pole in between)

```{r}
par(mar=c(1.5, 1.5, 0.5, 0.5))
mapPlot(coastlineWorld, longitudelim=c(-130, 50), latitudelim=c(60, 120),
       projection="+proj=stere +lat_0=90", col='gray')
```

This Arctic example also illustrates the specification of extra arguments in
the `projection` value, in this case the base latitude being set to the north
pole. See the table in the documentation for [`mapPlot`](../help/mapPlot) for a
listing of arguments, and for citations to external resources that explain what
they mean.

**Exercise 2.** Draw a view of Antarctica and the Southern Ocean.


## Mid-latitude views

The Lambert Conformal view is often used in maps that span wide longitudal
ranges, e.g. a map of Canada may be produced as follows.

```{r}
par(mar=c(1.5, 1.5, 0.5, 0.5))
mapPlot(coastlineWorld,
         longitudelim=c(-130,-55), latitudelim=c(35, 60),
         projection="+proj=lcc +lat_0=30 +lat_1=60 +lon_0=-100", col='gray')
```

**Exercise 3.** Plot the eastern North Atlantic using a Mercator projection.

**Exercise 4.** Plot the eastern North Atlantic using an Albers Equal-Area
Conic projection.



# Adding to map plots

Several functions are provided by oce to draw additional features on maps.
These include
[`mapText`](../help/mapText) for adding text,
[`mapPoints`](../help/mapPoints) for adding points,
[`mapLines`](../help/mapLines) for adding lines,
[`mapContour`](../help/mapContour) for adding contours, and
[`mapImage`](../help/mapImage) for adding images.
Each of these requires a map to have been drawn first with
[`mapPlot`](../help/mapPlot). The interfaces ought to be similar
enough to base-R functions that readers can decide how to accomplish 
common tasks.

**Exercise 5.** Plot the world in Molleweide projection, superimposing sea
surface temperature as contours.



# Solutions to exercises

**Exercise 1.** Plot the world with a Robinson projection, using gray shading
for land.

```{r fig.height=3}
par(mar=rep(0.5, 4))
mapPlot(coastlineWorld, col="lightgray", projection="+proj=robin")
```

**Exercise 2.** Draw a view of Antarctica and the Southern Ocean.
```{r}
par(mar=c(1.5, 1.5, 0.5, 0.5))
mapPlot(coastlineWorld, longitudelim=c(-180, 180), latitudelim=c(-130, -50),
       projection="+proj=stere +lat_0=-90", col='gray', grid=15)
```


**Exercise 3.** Plot the eastern North Atlantic using a Mercator projection.

```{r}
par(mar=c(1.5, 1.5, 0.5, 0.5))
mapPlot(coastlineWorldFine, col="lightgray",
        projection="+proj=merc",
        longitudelim=c(-80, -40), latitudelim=c(20, 60))
```

**Exercise 4.** Plot the eastern North Atlantic using an Albers Equal-Area
Conic projection.

In the solution given below, note how the `lon_0` argument is set to be at the
central value of `longitudelim`. This makes the 60W line be vertical on the
plot. The settings for `lat_1` and `lat_2` are the locations at which the cone
intersects the earth, so distortion is minimized at those latitudes. Note how
the lines of constant longitude are changed, compared with those in the
Mercator view.
```{r}
par(mar=c(1.5, 1.5, 0.5, 0.5))
mapPlot(coastlineWorldFine, col="lightgray",
        projection="+proj=aea +lat_1=30 +lat_2=50 +lon_0=-60",
        longitudelim=c(-80, -40), latitudelim=c(20, 60))
```

**Exercise 5.** Plot the world in Molleweide projection, superimposing sea
surface temperature as contours.

```{r fig.height=3}
par(mar=c(1.5, 1.5, 0.5, 0.5))
data(levitus, package="ocedata")
mapPlot(coastlineWorld, col="lightgray")
mapContour(levitus[['longitude']], levitus[['latitude']], levitus[['SST']])
```


# References

PROJ. PROJ.4 6.0.0 Documentation.” Accessed March 7, 2019.  https://proj4.org/.

Snyder, John P. Map Projections—a Working Manual. Washington: U.S. Geological
survey professional paper 1395, 1987. https://pubs.usgs.gov/pp/1395/report.pdf.

Snyder, John Parr. Flattening the Earth: Two Thousand Years of Map Projections.
Chicago, IL: University of Chicago Press, 1993.
https://www.press.uchicago.edu/ucp/books/book/chicago/F/bo3632853.html.

Snyder, John P., and Philip M. Voxland. “An Album of Map Projections.” U. S.
Geological Survey Professional Paper. Washington, DC, USA: U. S. Department of
the Interior: U. S. Geological Survey, 1994.
https://pubs.er.usgs.gov/publication/pp1453.


