---
title: "Subclassing oce objects"
author: "Dan Kelley (https://orcid.org/0000-0001-7808-5911)"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    fig_caption: yes
    fig_width: 5
    fig_height: 5
    dpi: 72
    dev.args: list(pointsize=11)
vignette: >
  %\VignetteIndexEntry{Subclassing oce objects}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

<!-- HOW TO BUILD THE VIGNETTE. -->
<!-- 1. edit this file in vignettes/oce.Rmd -->
<!-- 2. in an R console, use setwd() or otherwise move to the oce -->
<!--    directory, and then execute -->
<!--      devtools::build_vignettes() -->
<!-- 3. build oce. In the shell, you might use -->
<!--      cd oce ; echo "devtools::document(roclets=c('rd','collate','vignette'))" | R --no-save -->
<!--      CMD build --compact-vignettes="qpdf" oce -->
<!--      R CMD check oce_1.0-2.1.tar.gz -->
<!--      R CMD INSTALL oce_1.0-2.1.tar.gz -->
<!--    where the version number might need updating, or in an R console you might use -->
<!--      devtools::install(build_vignettes=TRUE) -->



```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

**Abstract.** This vignette explains how new classes of objects can be created,
using oce objects as a base class. The advantage of this is that the
newly-formed objects will automatically have important properties of oce
objects, in terms of operators such as `[[` and `[[<-`, functions
such as `subset()`  and `summary()`, schemes for handling units, etc.  The
treatment centres on the creation of a hypothetical class called `wave`, which
will hold time-series of elevation data.

# Tutorial

## Defining a class

The `setClass` function is the key to defining a new class. Entering
`?setClass` in an R console reveals the details of this function, including
some notes on how it has evolved since R version 3.0.

We need only a simple form here, with
```{r}
library(oce)
wave <- setClass(Class="wave", contains="oce")
```
being enough to create a new class called `wave` that inherits the base
features of the `oce` class.

To create a new object that inherits from the `wave` class, use e.g.
```{r}
w <- new("wave")
```

## Examining an object

We can see that `w` inherits from the `oce` class with
```{r}
class(w)
```
and a check that is common to see in code is
```{r}
inherits(w, "wave")
```

The contents of the object are revealed with
```{r}
str(w)
```

Notice that `w` three `oce` "slots", named `metadata`, `data` and
`processingLog`. These are inherited from `oce`.  The first is meant to hold
information about the data, such as a file-name, an instrument number, a
location of sampling, etc. The second is meant to hold actual data or
measurements.  And the third, not normally accessed by the user directly, holds
information about the object's evolution (note that the object considers itself
an `oce` object). The names of these three slots are usually enough to keep
them straight in the analyst's head, although it can sometimes be difficult
deciding whether something belongs in `metadata` or `data`.

The `oce` system also provides its objects with certain operators and
functions. For example, `[[` can be used to retrieve the slots, items within
the slots, and (in some cases) values that may be calculated from the contents
of the slots (e.g. `ctd` and similar objects related to hydrography can return
calculated potential temperature or Conservative Temperature, even though
neither is typically stored in such datasets).

That's not the only thing. All `oce` objects give special powers to the `[[`
operator, e.g. we can retrieve the `metadata` slot with
```{r}
w[["metadata"]]
```
and the same for the data slot
```{r}
w[["data"]]
```

## Adding data to objects

The `[[<-` operator can be used to fill the slots with information, e.g.  we
could insert a station location of `"STN01"` with
```{r}
w[["metadata"]]$station <- "STN01"
```
and verify that this worked with
```{r}
str(w)
```

However, there is a better way to insert metadata, with the `oceSetMetadata`
function, e.g.
```{r}
w <- oceSetMetadata(w, "serialNumber", 1234)
```
sets the serial number to 1234, and
```{r}
str(w[["metadata"]])
```
verifies that this worked.

Now, let's insert some data.  Imagine a half-minute dataset with 10Hz sampling,
for a signal with a $1$m elevation wave with period $10$s, plus some noise
of order $1$cm.
```{r}
t <- as.POSIXct("2019-01-01 00:00:00", tz="UTC") + seq(0, 30, length.out=100)
tau <- 10
e <- sin(as.numeric(2 * pi * as.numeric(t) / tau)) + rnorm(t, sd=0.01)
```
(Notice that we are not using the R `ts` form to make this time-series.  This
is because oceanographic data are commonly acquired on an irregular time
interval, so it makes sense to store observation time explicitly, instead of
using the start/step/stop approach of the `ts` scheme.)

These data may be inserted into our object with
```{r}
w <- oceSetData(w, "time", t)
w <- oceSetData(w, "elevation", e)
```

At this point the reader is likely to use `str` to see if this worked, but
since the object is starting to fill up, it might make sense to use the
`summary` function, which is inherited from `oce`.

```{r}
summary(w)
```

This produces a useful summary not just of the data, but also of how the object
was constructed. But we can do better. In some crazy world, someone might
consider measuring elevation in feet, not metres, and so we ought to specify
the unit. The way to do this is with the `unit` argument of `oceSetData`. This
is a somewhat tricky argument, as a study of the result of `?oceSetData` will
reveal. For now, we just show a common way, without explanation, writing
```{r}
w <- oceSetData(w, "elevation", e, unit=list(unit=expression(m),scale=""))
```

This over-rides the existing definition. Now, let's look at the summary:
```{r}
summary(w)
```
Notice that we now have a unit on the elevation, but we \emph{also} have an
indication that the value of that quantity was defined twice. This
processing-log feature is one of the big advantages of using `oceSetData`
over direct insertion into an object.

## Adding specialized functions

The most common function to add is a plot function. Since `plot` is a built-in
function, we are subclassing it.  The details of doing this are provided by
`?setMethod`. Again, studying the documentation for that function would be
worthwhile, but the gist is provided by a simple example, e.g.
```{r}
setMethod(f="plot",
    signature=signature("wave"),
    definition=function(x, which=1, ...) {
        if (which == 1) {
            plot(x[["time"]], x[["elevation"]], ...)
        } else if (which == 2) {
            hist(x[["elevation"]], ...)
        } else {
            stop("which must be 1 or 2")
        }
    })
```
Here, the `signature` argument tells R that `plot()` called with a `wave`
object as its first argument ought to use the indicated function.  That
function takes just two arguments: the object to be plotted, and `which`, an
indication of the desired plot type.

For example, since `which` defaults to 1, we can get a popular plot with
```{r}
plot(w)
```

Note the simplicity of this action. The user has no reason to state what kind
of object this is, because R detects the type and dispatches to the specialized
`wave`-plotting function. This may seem like a small thing in the present
context, but imagine an analyst writing code to analyse a wide variety of data
types: it is very convenient to have a simple function call that works for
each.

Since the `...` argument is passed into both the plotting methods. Thus, for
example, a cleaner time-series plot might be created with
```{r}
plot(w, type="l", xlab="Time [s]", ylab="Elevation [m]")
```

# Advanced methods

## Initializing objects with data

The following lets the user specify `time` and `elevation` when the object is
created. It also permits a specification of units, with a default being to us
metres.

```{r}
setMethod(f="initialize",
    signature="wave",
    definition=function(.Object, time, elevation, units) {
        if (missing(units)) {
            .Object@metadata$units <- list()
            if (missing(units))
                .Object@metadata$units$elevation <- list(unit=expression(m), scale="")
        }
        .Object@data$time <- if (missing(time)) NULL else time
        .Object@data$elevation <- if (missing(elevation)) NULL else elevation
        .Object@processingLog$time <- presentTime()
        .Object@processingLog$value <- "create 'wave' object"
        return(.Object)
    }
)
```


A test proves that this works as hoped for.
```{r}
ww <- new("wave", time=t, elevation=e)
summary(ww)
```
Notice that the units now appear, without complication to the user. Oh, and
this object now knows that it is a `wave` object, not an `oce` object \ldots
our little class is getting smarter by the minute!

## Specializing the `[[` operator

As in the previous section, the key of specializing how `[[` works is to use
`setMethod()`, but this time the function is named `"[["`. Suppose we want to
permit e.g. `w[["peak"]]` as a way to find the maximum value of wave height.
This becomes a call to the `"[["` function, with first argument as `w`, and
second argument as `"peak"`. We can handle this with:
```{r}
setMethod(f="[[",
    signature(x="wave", i="ANY", j="ANY"),
    definition=function(x, i, j, ...) {
        if (i == "peak") {
           return(max(x[["elevation"]], na.rm=TRUE))
        } else {
           callNextMethod()
        }
    }
)
```

(The details of the `signature` definition are explained in the documentation
provided by `?setMethod`, and readers ought to study that material before
changing the `signature` definition.)

The important thing to focus on is the `if` block in the function definition.
Called as e.g. `w[["peak"]]` causes `i` to equal `"peak"`, and so the return
value will be the maximum elevation. However, in all other instances, the
return values is provided by `callNextMethod()`, and what that does is to dig
one level deeper for a way to handle `[[`. At that deeper level, it finds the
`oce` definition, the details of which can be found with `?"[[,oce-method"`.

The test
```{r}
w[["peak"]]
str(w[["elevation"]])
```
verifies that our new code works for getting the peak value, and that it falls
back to the `oce` code for other calls.



