---
title: Notes on reading AD2CP 0x23 (echosounderRaw) data chunks
author: Dan Kelley
date: "`r Sys.Date()`"
---

**Abstract.** This is a diary of Dan Kelley's work, trying to add support for
the 0x23 (raw echosounder) AD2CP data type to the oce R package. To make things
concrete, I am showing my (R) code; even somebody not familiar with R ought to
get the gist.

## Prior to 2022-08-22

A document I received from Nortek in August of 2022 led me to believe that the
data were stored as pairs of `float32` entries^[Actually, the documentation
just says `float` but the entries are 8 bytes long, so I am assuming
`float32`.], with the first entry of each pair being echosounder amplitude, and
the second being echosounder quality.

However, this assumption yields some very odd results (lots of NA values, lots
of values that make no sense, like exponents in the 20s for quality).  I
communicated again, and was very please to get a quick and detailed reply
suggesting another format (see next).

## 2022-08-22

I am basing this on the contents of a Nortek email dated 22 Aug 2022
21:33:03.7921 (UTC).  I do not know how private that was, so I will not quote
from it in this document, nor name the sender.

The email told me that the data that oce calls `echosounderRaw` come in pairs
that represent complex numbers, with a `frac32` value for the real part and
then a `frac32` value for the imaginary part.

So, my next task was to learn what a `frac32` number is, since I have not come
across this notation.  I learned that it is used in the audio industry, so that
makes some sense in this context.

But, since I will need to make R/oce decode this format -- it's not a standard
C thing -- I will start by exploring the data in a step-by-step fashion.

We start by reading the extract file that Nortek sent me.  (This is the first
0x23 entry from a file I sent them.)

```{r}
library("testthat")                    # used for 'expect' functions
file <- "a5.0c.23.10.EchosounderRaw.101135.hex"
filesize <- file.info(file)$size
buf <- readBin(file, "raw", filesize)
expect_equal(buf[1], as.raw(0xa5))     # check that this starts a record
headerLength <- as.integer(buf[2])
expect_equal(headerLength, 12)         # check that header holds 12 bytes
type <- buf[3]
expect_equal(type, as.raw(0x23))       # check that it is 0x23
```

Good. The fact that no error messages are reported indicates that the first 3
bytes of the header make sense.  We may proceed to s extract the header and the
data sections, and to check the `offsetOfData` against what I know to be true
(from inspection of the header that is contained in the longer, original file).

```{r}
header <- buf[1:12]
data <- buf[13:filesize]
expect_equal(data[1], as.raw(0x01))    # Version must be 1
offsetOfData <- as.integer(data[2])
expect_equal(offsetOfData, 240)        # for this particular file, anyway
```

With these preliminaries over, we may proceed to check whether the data,
starting at `offsetOfData`, look like frac32 entries.

I have not found it easy to locate a definition of `frac32` data types, but at
https://code.ihub.org.cn/projects/729/repository/revisions/master/entry/modules/core/src/softfloat.cpp
I found a C macro
```
define fracF32UI( a ) ((a) & 0x007FFFFF)
```
(I found similar macros at other websites.)

I think this is intended to convert a 32-bit entry into a `frac32`.  Why do I
think that?  Well, in IEEE-754, the first bit indicates sign, and the next 8
bits indicate exponent, with the remaining 23 bits being the fraction. Since
the masking operation given above sets the first 9 bits to zero, it does appear
to be a way to extract the fractional part from a 32-bit floating-point number
(if in IEEE-754 format).  If the reader is not familiar with reading hex as
bits, the following shows the bit structure of 0x00 and then 0x7F
```{r}
paste(rev(ifelse(1==rawToBits(as.raw(0x00)),1,0)),collapse="")
paste(rev(ifelse(1==rawToBits(as.raw(0x7F)),1,0)),collapse="")
```

Now, I still have things to determine, the most important of which regards
sign.  I am given to understand that the values can range from (roughly) -1 to
+1, so sign must be stored somewhere. If so, it might be stored as the first
bit.  Or, I suppose, it might be stored within the fractional part (that is, it
might be interpreted as an integer with a sign bit).

To make serious progress, I need to learn more from Nortek about this format.
However, I will explore a bit more, in the mean-time.

First, we extract the measurements.
```{r}
echosounder <- data[seq(offsetOfData+1, length(data))]
```
From other work (including reading the header from the file from which this
extract-data was derived) I know that there ought to be 1974 8-byte chunks in
`echosounder`, so we test on that before proceeding.
```{r}
expect_equal(length(echosounder)/8, 1974)
```

Good.  I have some confidence in having extracted the data, without
off-by-one-byte errors.

Now, if the data are in `frac32` format, then the first byte in each 4-byte
chunk ought to be 0x00 (that will be the result of masking, as above).  Well,
is that the case?  As a first guess, we try
```{r}
sum(echosounder==as.raw(0x00)) / length(echosounder)
```
and the result is a bit over 1/4, which is at least some confirmation.  But
a better test will be to look at the actual data.

To make it clearer, I'll make a table in 8 columns, so that column 1 through 4
will correspond to the real part, and columns 5 through 8 will correspond to
the imaginary part.

```{r}
m <- matrix(echosounder, byrow=TRUE, ncol=8)
head(m, 30)
```

This is unexpected.  If these were `frac32` data, with zero exponents, then
there should only be two possibilities for column 1, 0x00 (bits 00000000) and
0x80 (bits 10000000).  The same should be true of column 5. However, both
columns 1 and 5 have a multiplicity of values.

Although this is clutching at straws, I *do* see some patterns in the data,
e.g. columns 3 and 4, and also 7 and 8, fall into repeating patterns. But I did
not acquire these data myself, so I don't know what to make of these patterns.

**A question for Nortek**

My reading of the data suggests that they are not actually in `frac32` format.
We knew that from initial tests (reading as `float32` yield many values with very
large exponents, which is inconsistent with zero exponent, of course).

**Question for Nortek:** Are the data stored in `float32` that ought to be
converted to `frac32` prior to use?  (In other words, is the idea that those
nonzero exponents are to be discarded?)

**Comment to Nortek:** The data-format story has been evolving in the past few
weeks, and I really appreciate your patience in helping me to understand the
format, so that R users can work with their data within a familiar analysis
environment.


## 2022-08-23

A Nortek email dated 23 Aug 2022 09:00:00.5182 (UTC) tells me that the idea is
to read as 32-bit signed integer, then convert to signed 32-bit float, then
divide by 2^31.  Oce now does that (commit
daaddc05ba8a85cb65b44a39e9d2c70ba0aa696a of the"develop" branch on
github.com/dankelley/oce), and below is what it yields (please ignore the
debugging output from `read.oce()`, which is temporary).

The numbers do not tell me much, since I have no independent way to read the
data, but at least the plot seems to suggest that the data are not fully random.


```{r}
library(oce) # github.com/dankelley/oce "develop" branch
fn <- "a5.0c.23.10.EchosounderRaw.101135.hex"
dn <- read.oce(fn)
mod <- Mod(dn@data$echosounderRaw$samples)
mod[1, 1:10]
par(mfrow=c(2,2))
plot(mod[1,], type="s")
plot(log10(mod[1,]), type="s")
# perhaps a plot will be useful to summarize
hist(log10(mod), breaks=1000, main="")
imagep(log10(mod))
```

Although I did not acquire these results myself, the plots have a character
that is suggestive of echosounder data, so I think we are on the right track.

**Question for Nortek:** Does the above seem reasonable to you? Do you get the
same values for the first few real and imaginary values?

**Comment to Nortek:** Things are starting to look good to me, and I have to
thank you for this!

## Addendum (a longer private file)

The results shown below look reasonable, I think.  Note that I have not stored
time and distance into `@data$echosounderRaw` yet ... my first task was to read
the data, which has proved tricky, owing to the fact that Nortek documents are
still being written for this data type.

```{r}
library(oce) # github.com/dankelley/oce "develop" branch
par(mfrow=c(1,1))
file <- "~/Dropbox/oce_secret_data/ad2cp/s100_truncated.ad2cp"
if (file.exists(file)) {
    d <- read.oce(file)
    imagep(log10(Mod(d@data$echosounderRaw$samples)))
}
```


