---
title: dek05 notes on reading AD2CP 0x23 (echosounderRaw) data chunks
author: Dan Kelley
date: "`r Sys.Date()`"
---

**Abstract.** This is a diary of Dan Kelley's work, trying to add support for
the 0x23 (raw echosounder) AD2CP data type to the oce R package. To make things
concrete, I am showing my (R) code; even somebody not familiar with R ought to
get the gist.

```{r echo=FALSE}
knitr::opts_chunk$set(dev="png", dev.args=list(type="cairo-png"))
```

# Diary of updates

## Update 2022-09-01

Ragnar supplied a formula for `sampleStartIndex`, as well as a very helpful
diagram explaining `XMIT`.  From these things I was able to infer a formula for
`echosounderRaw$cellSize`; this formula yields the graphs provided near the end
of the present document.

## Update 2022-08-29

I've learned that what I call `blankingDistanceInCm` is incorrectly stored in
AD2CP files. In fact, it is always FALSE (i.e. the factor is 1e-3, not 1e-2).
This is detected in `read.adp.ad2cp()` now, and a warning is issued if the
value stored in the file does not yield FALSE.  With this scheme, the warning
will not appear when Nortek updates their systems and folks generate files with
the newer system.


## 2022-08-22

I am basing this on the contents of a Nortek email dated 22 Aug 2022
21:33:03.7921 (UTC).  I do not know how private that was, so I will not quote
from it in this document, nor name the sender.

The email told me that the data that oce calls `echosounderRaw` come in pairs
that represent complex numbers, with a `frac32` value for the real part and
then a `frac32` value for the imaginary part.  At first I though this meant a
format mentioned at
https://code.ihub.org.cn/projects/729/repository/revisions/master/entry/modules/core/src/softfloat.cpp
but after some email exchanges, I found that it referred to a scaled signed
integer.  The recipe is simple: read as 32-bit signed integer, convert to numeric, then divide by 2^31.


## Prior to 2022-08-22

A document I received from Nortek in August of 2022 led me to believe that the
data were stored as pairs of `float32` entries^[Actually, the documentation
just says `float` but the entries are 8 bytes long, so I am assuming
`float32`.], with the first entry of each pair being echosounder amplitude, and
the second being echosounder quality.

However, this assumption yields some very odd results (lots of NA values, lots
of values that make no sense, like exponents in the 20s for quality).  I
communicated again, and was very pleased to get a quick and detailed reply
suggesting another format (see previous).
# Latest results

A test file (from which `tests/testthat/local-data/ad2cp/ad2cp-01.ad2cp` is
extracted) produces the following results.  Note that this uses the
`startSampleIndex` value that is stored in the file, which is an integer, and
not a value calculated using Ragnar's formula.

```{r fig.width=8, fig.height=8, dev.arg=list(pointsize=10)}
library(oce) # github.com/dankelley/oce "develop" branch
file <- "~/Dropbox/oce_secret_data/ad2cp/s100_truncated.ad2cp"
layout(rbind(c(1:2),c(3:4)), width=c(2/3,1/3))
if (file.exists(file)) {
    d <- read.oce(file)
    time <- d@data$echosounder$time
    distance <- d@data$echosounder$distance
    e <- log10(d@data$echosounder$echosounder)
    imagep(time, distance, e, quiet=TRUE, ylab=resizableLabel("distance"))
    ylim <- par("usr")[3:4]
    em <- apply(e, 2, mean)
    wem <- which.max(em)
    par(mgp=c(2,0.7,0))
    plot(em, distance,
        xlab="intensity", ylab=resizableLabel("distance"),
        type="l", ylim=ylim, yaxs="i")
    mtext(sprintf("e peak @ %.0fm", distance[wem]))

    rtime <- d@data$echosounderRaw$time
    rdistance <- d@data$echosounderRaw$distance
    re <- log10(Mod(d@data$echosounderRaw$samples))
    imagep(rtime, rdistance, re, quiet=TRUE, ylab=resizableLabel("distance"))
    ylim <- par("usr")[3:4]
    rem <- apply(re, 2, mean)
    # the max is in first samples (during XMIT burst, I think)
    wrem <- 20 + which.max(tail(rem, -20))
    par(mgp=c(2,0.7,0))
    plot(rem, rdistance,
        xlab="intensity", ylab=resizableLabel("distance"),
        type="l", ylim=ylim, yaxs="i")
    mtext(sprintf("eraw peak @ %.0fm", rdistance[wrem]))
}
```

Two questions remain:

1. Should we use the integer `startSampleIndex` that is in the file, or should
   we compute it using the formula provided by Ragnar?  The former is an
   integer value that is 16 in a sample file, and if that's typical then
   rounding might be expected to give about 3% error in the results for
   `cellSize` and thus `distance`. The graph above uses the integer value. If
   the calculated `startSampleIndex` were used instead, the peak would shift
   from 282m to `r round(with(d@data$echosounderRaw,cellSize2*282/cellSize))` m.

2. What `soundSpeed` should be used in computations of `cellSize`?  In the
   graph presented above, a mean of `soundSpeed` values from the file was used.

