---
title: dek06 computing distance for 0x23 (echosounderRaw) data chunks
author: Dan Kelley
date: "`r Sys.Date()`"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

**Abstract.** Ellie Best's (2022-08-28) formula for the `cellSize` of
`echosounderRaw` (0x23) samples yields a distance range that is 29 percent
smaller than that of `echosounder` (0x1c). I speculate here on a replacement
formula, but I'm basically clutching at straws, and hope someone at Nortek can
comment on this.

In the oce code, echosounder distance vectors are computed from variables named
`blankingDistance`, `cellSize` and `numberOfSamples`.  Unfortunately, this method cannot
be applied to `echosounderRaw`, since its data stream does not store
`cellSize`.  A solution was suggested in EB's email dated on or near
2022-08-28: compute `echosounderRaw$cellSize` as the ratio of
`echosounder$blankingDistance` to `echosounderRaw$startSampleIndex`.
(Here and henceforth, I'll use R notation sometimes; the dollar sign takes the place of a period in some other languages.)

```{r results="hide", echo=FALSE}
suppressPackageStartupMessages(library(oce))
d <- read.oce("~/git/oce/tests/testthat/local_data/ad2cp/ad2cp_01.ad2cp")
```

When I do this with the sample file I call `ad2cp_01.ad2cp` that we
have been discussing for a few weeks now, I get maximal `echosounderRaw` (0x23)
distance
`r max(d@data$echosounderRaw$distance)`
m, which is at odds to the
`echosounder` maximal distance of
`r max(d@data$echosounder$distance)`
m.
(For context, this 29 percent difference greatly exceeds the 6 percent difference that would result by
changing the value of `startSampleIndex` by 1.)

To see this (running the very latest github.com/dankelley/oce "ad2cp" source),
do as follows.

```{r eval=FALSE}
library(oce)
d <- read.oce("~/git/oce/tests/testthat/local_data/ad2cp/ad2cp_01.ad2cp")
```


```{r}
max(d@data$echosounderRaw$distance)
max(d@data$echosounder$distance)
```

Sensing something was wrong, I worked in steps to explore the computation.

First, I verified in various ways that I am inferring
`echosounderRaw$numberOfSamples` correctly as
`r d@data$echosounderRaw$numberOfSamples`.  Since the maximal
distance is the product of this distance and the `echosounderRaw` cell size,
I'll turn my focus to the computation of that cell size.

The code is presently inferring `echosounderRaw$cellSize` using EB's formula, that is,
dividing `echosounder$blankingDistance`
(`r d@data$echosounder$blankingDistance` m) by a quantity I name as
`echosounderRaw$startSampleIndex`
(`r d@data$echosounderRaw$startSampleIndex`).  This yields
`r with(d@data, echosounder$blankingDistance/echosounderRaw$startSampleIndex)` m.
Multiplying this by `echosounderRaw$numberOfSamples`
(`r d@data$echosounderRaw$numberOfSamples`) yields
maximal distance `r max(d@data$echosounderRaw$distance)` m as stated above.

I scanned the documentation for clues, but did not find anything.  Then I
looked at the data header, and noticed something called `RANGESTART1` in the
`READECHO` header line. We can retrieve this value with
```{r}
ad2cpHeaderValue(d, "READECHO", "RANGESTART1")
```
Perhaps -- and I admit to clutching at straws -- we ought
to take this as an origin offset.  Trying this with
```{r}
BDnew <- d@data$echosounder$blankingDistance - ad2cpHeaderValue(d,"READECHO","RANGESTART1")
CSnew <- BDnew / d@data$echosounderRaw$startSampleIndex
maxDistance <- CSnew * d@data$echosounderRaw$numberOfSamples
```
yields `maxDistance`=`r maxDistance` m, which indeed is quite close to the
maximal distance of `r max(d@data$echosounder$distance)` m computed
for the `echosounder` data stream.

**Hypothesis** The suggested formula for `echosounderRaw$cellSize` as the ratio
of `echosounder$blankingDistance` to `echosounderRaw$startSampleIndex`, should
be amended by subtracting `RANGESTART1` from the numerator.

**Question for Nortek:** Is there anything sensible in the above?  Basically,
is there merit to my modification of EB's formula, or is the improved agreement
just a coincidence?  And, if there *is* merit to my idea, is there a way I can
infer `RANGESTART1` from the data stream, without looking at the header?  (I
ask about the header-less case because oceanographers sometimes break big files
up into pieces, and only the first would have the header.)



