---
title: "oce: an R package for Oceanographic Analysis"
tags:
  - R
  - oceanography
  - measurement
  - data analysis
authors:
  - name: Dan E. Kelley^[Corresponding author.]
    orcid: 0000-0001-7808-5911
    affiliation: 1
  - name: Clark Richards
    orcid: 0000-0002-7833-206X
    affiliation: 2
  - name: Chantelle Layton
    orcid: 0000-0002-3199-5763
    affiliation: 3
affiliations:
 - name: Dan E. Kelley, Professor, Dalhousie University
   index: 1
 - name: Clark Richards, Research Scientist, Bedford Institute of Oceanography, Department of Fisheries and Oceans, Canada; also Adjunct Professor, Dalhousie University
   index: 2
 - name: Chantelle Layton, Physical Scientist, Bedford Institute of Oceanography, Department of Fisheries and Oceans, Canada
   index: 3
date: 26 May 2021
year: 2021
bibliography: paper.bib
output: rticles::joss_article
csl: apa.csl
journal: JOSS
---

# Statement of Need

Oceanographic field experiments often employ a suite of instrument types, each
reporting data in a different format. Many of these formats are complex and
difficult to decode. Although manufacturers usually provide software for
accessing data produced by their instruments, it tends to be proprietary and
closed-source. This is a problem for researchers seeking to analyse their data
in novel ways, or combine data from multiple instruments. The `oce` package
[@kelley__aut_oce_2021] addresses such issues by providing functions that
handle dozens of data formats. In addition, it facilitates specialized
calculations and data displays that are particular to the discipline.  Since
`oce` is written in the R language
[@ihaka_r_1996;@r_core_team_introduction_2021], it forms a link to an array of
more general tools that oceanographers may need in their work
[@kelley_oceanographic_2018].

# Overview

The `oce` package has been hosted on CRAN [@noauthor_comprehensive_2021] since
the year 2009. The CRAN version, which is updated once or twice a year, may be
installed by typing `install.packages("oce")` in an R console. Users who need
newer features may use `remotes::install_github("dankelley/oce",ref="develop")`
to download and build the development branch. Those wishing to view or
participate in the development process are welcome to do so, at
\url{https://github.com/dankelley/oce}.

The package has functions for decoding many data formats.  These functions
return S4 objects with slots holding (a) the data, (b) related metadata, and
(c) a log of `oce` functions that made the object.  This is illustrated by
executing the following in an R session, for a built-in object creating by
reading a profiling instrument called a CTD.
```{r setup, include=FALSE}
library(oce)
options(width=70)
knitr::opts_chunk$set(fig.path="", dev="png", dpi=300, pointsize=8)
data(ctd)
```
```{r eval=FALSE}
library(oce)                           # load library
data(ctd)                              # load a built-in sample file
slotNames(ctd)                         # see 'slot' names
```

The next step after loading an object, or reading it from a data file, is often
to get a textual overview with `summary(ctd)`, or a graphical overview, e.g. with
`plot(ctd)` producing Figure 1.

```{r echo=FALSE,results="hide", fig.cap="CTD analysis example. Top left: depth variation of temperature and salinity. This uses pressure as a surrogate for depth, with 1 decibar corresponding to approximately 1 metre of depth. Top right: variation of density excess over 1000 kg/m$^3$, and $N^2$, a measure of the gravitational stability of the water column. Bottom left: co-variation of salinity and temperature. Bottom-right: map of the study region, with gray being land in a coarse representation of the coastline.", dev.args=list(pointsize=9)}
plot(ctd)
```

It is also common to exert fine-grained control of graphical representations,
with e.g.  `plot(ctd, which="temperature")` to plot just the temperature
variation with depth (results not shown here). The variations of other
properties may be shown by setting `which` appropriately, and this argument can
also be used to specify other types of plots, in addition to the
depth-variation form.

Besides this `"ctd"` subclass, `oce` supports dozens of other subclasses that
cover a wide range of oceanographic instrumentation.  In every case, the same
`"summary()"` and `"plot()"` function calls provide textual and graphical
representations of the data.  This specialization of these two generic
functions simplifies analysis considerably.  For example, if `PATTERN` is a
regular expression that specifies a set of data files, whether of a single
instrument type or multiple instrument types, then
```{r eval=FALSE}
for (file in list.files(PATTERN)) {
    d <- read.oce(file)
    summary(d)
    plot(d)
}
```
will provide information about each data file of interest, forming a good
first stage of analysis.

Oce also provides other generic functions, such as `subset()` for focusing on
subsets of data, `handleFlags()` for processing data-quality flags, and `[[`
for accessing data.

The accessor operation is particularly worthy of note, for several reasons, of
which two stand out.

1. `[[` lets users access information regardless of where it occurs in the
   object. For example, a CTD does not measure longitude and latitude, but
these things are often streamed into data files separately, so they are stored
in the object's `metadata` slot, not the `data` slot.  Users do not need to
know this detail, though, because e.g. `ctd[["longitude"]]` will extract
longitude from whichever slot holds it.  For other data types, e.g. from a GPS
instrument, the `[[` operation might get it from the `data` slot.  Importantly,
the same code fragment will would work for both data structures.  As above, the
benefit is that code written for one object type will work for another.

2. `[[` can access not just information stored within the object, but also
   things that can be calculated from that information. For example, CTD files
typically hold temperature along with other properties from which seawater
density may be computed [@millero_history_2010;@mcdougall_getting_2011].
Density is an important dynamical factor, and so `oce` is set up so that e.g.
`ctd[["density"]]` will compute it, provided that the named object contains the
required elements.  This also applies for other elements that are commonly
needed for oceanographic analysis, but not measured directly.

Importantly, `[[` acts as a bridge from the oceanographic realm to the general
R realm, letting users transform their data for further analysis.  With tens of
thousands of R packages being available, this lets oceanographers focus on
their work, not on tool generation.

# Example: Tidal Analysis

A more detailed example may help to solidify some of the key aspects of `oce`.
Many readers will have an interest in tides, so we will work with a year-long
record of sea level, $\eta=\eta(t)$ in Halifax Harbour, in the year 2003,
during which the city was struck by Hurricane Juan.

Consider the code given below, which produces Figure 2.  A built-in sealevel
file is used, to make a reproducible example, but replacing the `data()` call
with a `read.sealevel()` call will handle data files in standard formats.  Note
that the `tidem()` function is fairly sophisticated with over 500 lines of R
code being used to apply the specialized procedures of tidal analysis
[@godin_analysis_1972;@pawlowicz_classical_2002;@foreman_versatile_2009].
Readers who see that the function evokes the `lm()` function for linear models,
may not be surprised that `oce` provides a function named `predict()`, for
generating tidal predictions.

```{r results="hide", fig.cap="Tidal analysis example. Top: a year of sea level variation in Halifax Harbour. Bottom: sea level after removing a fit to the tides.", dev.args=list(pointsize=9)}
library(oce)                           # load library
data(sealevel)                         # use built-in example dataset
t <- sealevel[["time"]]                # extract time
eta <- sealevel[["elevation"]]         # extract sea level
m <- tidem(sealevel)                   # fit tidal model
etaDetided <- eta - predict(m)         # de-tide observations
par(mfrow=c(2, 1))                     # set up a two-panel plot
oce.plot.ts(t, eta, xaxs="i",          # top: observed sea level
    grid=TRUE, ylab="Sea level [m]")
oce.plot.ts(t, etaDetided, xaxs="i",   # bottom: de-tided sea level
    grid=TRUE, ylab="De-tided sea level [m]")
```

A comparison of the panels of Figure 2 reveals that tides explain much of the
sea level variation in Halifax Harbour.  The lower panel illustrates an
increase of detided variance during the winter months, as expected at a
northern mid-latitude.  More surprising is the large spike towards the end of
September.  This is a result of Hurricane Juan, which swept over Halifax at
that time, causing a storm surge of approximately 1.5m that, along with high
waves, caused major damage in the harbour [@xu_extreme_2012].  (Readers might
find it informative to supply an `xlim` argument to the plot calls, to narrow
in on the event.)

# Conclusions

The `oce` package provides for many aspects of oceanographic analysis, having
evolved in an open-source environment for more than a decade.  The developers
have benefited from a supportive user community, members of which have
contributed insightful bug reports and suggestions for improvements.  New
features are added continually, to handle new instrument types, new data
repositories, and new methods. Physical oceanography is a major focus of the
package, but we hope this paper will generate interest in other communities,
ranging from climatologists to those in marine disciplines such as chemistry
and biology.  Our other goal is to encourage the development of new R packages,
such as `argoFloats` [@kelley_argofloats_2021], that build upon `oce`.

# References

