---
title: Investigating an odd RDI file
author: Dan Kelley
date: "`r Sys.Date()`"
output: pdf_document
---

# Introduction

A user tried reading what I'm told is a Sentinel RDI file, but found that it
failed.  I have the (91 Mb) file locally, and this document is the
result of my investigation of its contents, after I confirmed that, indeed,
`read.adp.rdi()` cannot read the file.

# How does the file start?

According to RDI documentation in my possession (from a few months ago, see
References, and going back to which 2006), the first two bytes of an RDI file
must be 0x7f 0x7f. Using `od` to inspect the file,

```
od -x ~/Downloads/wetransfer_dplsp000-000_2023-03-20_0847/DPLSP000.000 | head -1
```
yields
```
0000000      797f    0056    0100    0008    0103    2832    41cc    6029
```
so, clearly, this file does not match expectation.

Note that the above is as shown on a little-endian machine, so the byte order
is 0x7f followed by 0x79.  

It is convenient to use R to examine the file.  The first 16 bytes  are as
follows (as expected, by transposing bytes in the `od` output).
```{r}
f <- "~/Downloads/wetransfer_dplsp000-000_2023-03-20_0847/DPLSP000.000"
buf <- readBin(f, "raw", 1e6) # read a million bytes
buf[1:16]
```

# Is the starting byte pair repeated?

Using
```{r}
l1 <- which(buf == as.raw(0x7f))
l2 <- which(buf == as.raw(0x79))
n <- min(length(l1), length(l2))
df <- data.frame(l1[1:n], l2[1:n])
colnames(df) <- c("0x7f", "0x79")
df[1:10, ]
```
reveals a pairing pattern, with interruptions.  (The interruptions are not
unexpected, since these values might occur within data, not just at the starts
of data chunks.)  If we are willing to skip some values, we can see pairs here,
at 1-2, then (skipping 73) at 89-90, then at 179-180, 269-270, 359-360,
449-450, and (skipping 529) 539-540, then at 629/630 and 719/720.

There is also a pattern to pair separation, with values 88, 90, 90, 90, 90, and
90 bytes.  Perhaps the first chunk is of one type, and the others are of
another type.


# Is this byte pattern repeated later in the file?

We may examine byte distribution (over the first Mb, say) with
```{r}
d1 <- diff(df[, 1])
d2 <- diff(df[, 2])
knownDiff <- c(90, 973) # by inspection
par(mfrow=c(2, 1), mar=c(2,3,1,1), mgp=c(2, 0.7, 0))
plot(d1, col=ifelse(d1 %in% knownDiff, 2, 4),
    xlab="", ylab="diff(index of 0x7f)", type="p", cex=0.3)
plot(d2, col=ifelse(d2 %in% knownDiff, 2, 4),
    xlab="", ylab="diff(index of 0x79)", type="p", cex=0.3)
```
In this graph, some values that seemed (by eye and histogram) to be common are
shown in red.  Again, a pattern is discernible, but it is not simple.  Both
byte values repeat frequently at an index separation of 90, but the longer
separation of 973 only occurs for 0x7f.  And, more oddly, the high-separation
0x79 values occur at several separations.

# Can oce read the file if the byte-pair is changed?

I have tried altering the existing oce code to recognize the start-pair 0x7f
0x79 but the values that get read do not make much sense. Frankly, this is not
especially surprising.  If I were an RDI engineer, and we had a chance to make
a whole new format, I might change things.  Perhaps new things need to be
stored now. Perhaps the bit-packing has run into limitations, requiring an
extra byte here and there. Perhaps the checksum algorithm would be more
error-tolerant if it were changed. Changes along these lines cannot be inferred
by looking at bytes and guessing.  We need a manual that documents the format,
and the Sentinel V manual from January 2023 (see References) is not such
a manual, because it explicitly indicates that the ensemble leading bytes must be 0x7f 0x7f.

So, for now, oce cannot read these files.

# References

Teledyne Marine. “Sentinel V Self-Contained Sentinel V Real-Time Operation Manual,” January 2023. http://www.teledynemarine.com/Documents/Brand%20Support/RD%20INSTRUMENTS/Technical%20Resources/Manuals%20and%20Guides/Sentinel%20V%20and%20Velocity/Sentinel_V_SC_RT_Operation_Manual.pdf.

